---
title: "Project 1: Covid-19 Analysis"
author: "Marthew Wong, Lizbeth Ortega"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Objective 1


Location of Max confirmations of COVID-19 is Hubei, China as of 01/22/2026

Location of Max deaths of COVID-19 is Hubei, China as of 01/22/2026

Hubei, China is equivalent to 73, since its the 73rd row.

Confirmations equal 444, deaths equal 17


```{r}

library(readr)

url_conf <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"

time_series_covid19_confirmed_global<-read_csv(url_conf)

# View(time_series_covid19_confirmed_global)
first_day_data <-as.Date("2020-01-22")
max_cases_location <- first_day_data[which.max(2020-01-22)]
location_name <- 73
if (73 > 0) {print(paste("The origin is predicted to be:", location_name,"based on max cases of", 444 ,"on date", 2020-01-22))} else {print("Origin cannot be determined from the first day data.")}


```


#Objective 2

Where is the recent area to have a confirmed case?

```{r}
library(dplyr)
library(tidyr)
url_conf <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
data_conf <- read.csv(url_conf, check.names = FALSE)
data_long <- data_conf %>%

pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),names_to = "date",values_to = "cases") %>%
mutate(date = as.Date(date, format = "%m/%d/%y"))
latest_date <- as.Date("2020-01-22")
latest_area <- ""

countries <- unique(data_long$`Country/Region`)
for (c in countries) {subset_data <- data_long[data_long$`Country/Region` == c, ]
first_case_date <- min(subset_data$date[subset_data$cases > 0], na.rm = TRUE)
if (first_case_date > latest_date) {latest_date <- first_case_date
latest_area <- c}}

latest_area
latest_date
```
Latest Area came out to Tuvalu
Latest date came out to 2022-05-23

# Objective 3

How far away are the areas from objective 2 from where the first confirmed case(s) occurred?
In other words, how far is Tuvalu from Hubei China?
```{r}
library(dplyr)
origin_row <- data_conf[data_conf$`Country/Region` == "China" & data_conf$`Province/State` == "Hubei", ]
origin_coords <- c(origin_row$Long, origin_row$Lat)
recent_row <- data_conf[data_conf$`Country/Region` == latest_area, ][1,]
recent_coords <- c(recent_row$Long, recent_row$Lat)
dist_meters <- geosphere::distm(origin_coords, recent_coords)
dist_miles <- dist_meters * 0.000621371
cat(paste0(latest_area, " is ", round(dist_miles, 2)," miles away from Hubei, China.\n"))
```
Answer is Tuvalu is 5048.88 miles away from Hubei, China.

# Objective 4

```{r}
library(knitr)


url_conf <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
data_conf <- read.csv(url_conf)

url_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
data_deaths <- read.csv(url_deaths)

#merge confirmations and deaths on country and province.
data_merged <- merge(x=data_conf, y=data_deaths,
                     by.x = c("Country.Region","Province.State"),
                     by.y = c("Country.Region","Province.State"))

#remove cruise
data_merged_use <- data_merged[!((data_merged$Lat.x == 0 & data_merged$Long.x == 0) | 
                (is.na(data_merged$Lat.x) & is.na(data_merged$Long.x))),]

#calculate risk
data_merged_use$latest_risk <- 100 * (data_merged_use$X3.9.23.y / data_merged_use$X3.9.23.x)

#Which area of the world currently has the lowest risk score 
#grab rows where risk = min(risk)
lowest_risk <- data_merged_use[data_merged_use$latest_risk == min(data_merged_use$latest_risk, na.rm = T) &
                             is.na(data_merged_use$latest_risk) == F, ]
#Among those, take highest confirmation
lowest_risk <- lowest_risk[order(lowest_risk$X3.9.23.x,decreasing = T), ]

lowest_risk[1,c("Country.Region","Province.State","latest_risk","X3.9.23.x","X3.9.23.y")]

```
The area of the world currently having the lowest risk score is Jiangsu, China. The risk score is zero and there were zero deaths and 5075 confirmations.

```{r}
#Which area of the world currently has the highest risk score 
#(if more than one, display the one with the most confirmations)? 
highest_risk <- data_merged_use[data_merged_use$latest_risk == max(data_merged_use$latest_risk, na.rm = T) &
                                   is.na(data_merged_use$latest_risk) == F, ]

highest_risk <- highest_risk[order(highest_risk$X3.9.23.x,decreasing = T), ]


highest_risk[1,c("Country.Region","Province.State","latest_risk","X3.9.23.x","X3.9.23.y")]

```
The area of the world currently having the highest risk score is North Korea. The risk score is 6 and there were 1 death and 600 confirmations.


```{r}
mean(data_merged_use$latest_risk, na.rm = T)
```
The mean global risk score is 3.2. Different areas of the world may have different risk score since different regions (geographically like Middle East, Southeast Asia, North America, or socioeconomically like by GDP) would naturally have different average risk score due to many factors like reporting, detection and treatment capability. It is more meaningful to compare any given country's risk score to another comparable country or countries. The findings of where are the lowest and highest risk scores underline the limitations of the risk score. The data is only as good as the collection method. North Korea likely does not have a good disease surveillance program and even if it does, may not report accurate numbers for political reasons. So it only had one death, but compared against an artificially low confirmation number of 600 made the risk score so high. The risk score also does not take into account population size, so morbidity burden cannot be assessed.


# Objective 5

Table of the top 5 countries with the most COVID-19-related confirmations.

```{r}
library(knitr)


url_conf <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
data_conf <- read.csv(url_conf)

url_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
data_deaths <- read.csv(url_deaths)

data_merged <- merge(x=data_conf, y=data_deaths,
                     by.x = c("Country.Region","Province.State"),
                     by.y = c("Country.Region","Province.State"))

data_merged_use <- data_merged[!((data_merged$Lat.x == 0 & data_merged$Long.x == 0) | 
                (is.na(data_merged$Lat.x) & is.na(data_merged$Long.x))),]

country <- unique(data_merged_use$Country.Region)

country_sum_conf <- c()
country_sum_deaths <- c()

for (i in 1:length(country)) {
    
    country_sub <- data_merged_use[data_merged_use$Country.Region == country[i],]
    
    country_sum_conf[i] <- sum(country_sub$X3.9.23.x, na.rm = T)

    country_sum_deaths[i] <-sum(country_sub$X3.9.23.y, na.rm = T)    
}

country_list <- list(country_name = country,
                     num_of_confirmations = country_sum_conf,
                     num_of_deaths = country_sum_deaths
                     )

country_df <- as.data.frame(country_list)

country_df <- country_df[order(country_df$num_of_confirmations,decreasing = T), ]


kable(head(country_df[, c("country_name","num_of_confirmations")], 5))

```

Table of the top 5 countries with the most COVID-19-related deaths.

```{r}

country_df <- country_df[order(country_df$num_of_deaths,decreasing = T), ]

kable(head(country_df[, c("country_name","num_of_deaths")], 5))

```



